Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-05-19 LICS_2025/monoidal_e_graphs/sections/appendix.tex
wip
camera-ready and arxiv script

\subsection{$\catname{SLat}$}
\label{sec:appendix:slat}

@@

    The unit for this tensor product is $I = \{*\}$ --- a one-element semilattice.
    Clearly $S \otimes I \cong S$ by mapping $(s,*) \mapsto s$ for all $s \in S$ and vice versa.
    The symmetry, associators and unitors are then obvious morphisms.
    Finally, the category is closed since the set of homomorpisms between two semilattices is a semilattice by defining $(f + g)(x)$ as $f(x) + g(x)$.
    $f + g$ is a homomorphism since $(f + g)(x+y) = f(x+y) + g(x+y) = f(x) + f(y) + g(x) + g(y) = f(x) + g(x) + f(y) + g(y) = f(x+y) + g(x+y)$.
  \end{proof}

  This makes $\catname{SLat}$ a suitable base for enrichment and allows us to take $\mathcal{V} = \catname{SLat}$.

@@

    \begin{proposition}[Special case of Proposition 6.4.6~\cite{Borceux_1994}]
      The forgetful functor $U : \catname{SLat} \to \catname{Set}$ has a left adjoint free functor $F : \catname{Set} \to \catname{SLat}$.
    \end{proposition}
    \begin{proof}
      The functor $F$ is defined by letting $F(A) = \coprod_{A} I_{\catname{SLat}}$ where the latter is a coproduct which is a `free' product of semilattices defined as follows.
      The elements of $S_{1} \coprod S_{2}$ are sequences $s_{1} + s_{2} + \ldots + s_{n}$ where each $s_{i}$ is either from $S_{1}$ or $S_{2}$ quotiented by all relations in $S_{1}$ and $S_{2}$.
      The adjunction is then given by the following natural isomorphisms
      \begin{align*}
      \catname{SLat}(\coprod_{A}(I_{\catname{SLat}}), B) &\cong \prod_{A}(\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A,\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A, U(B))
      \end{align*}
      The fist isomorphism is given by the fact that hom-functor makes limits into colimits in its first argument, the second being given by the property that $|\catname{Set}(A,B)| = |B|^{|A|} = |\underbrace{B \times \ldots \times B}_{|A|}|$, and the last isomorphism is given by the definition of $U$.
      Furthermore, we have, 
      \[
      F(I) \cong I_{\catname{SLat}}
      \]
      and 
      \begin{align*}
      F(A) \otimes F(B) &\cong (\coprod_{A} I_{\catname{SLat}}) \otimes (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (I_{\catname{SLat}} \otimes \coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (\coprod_{B} (I_{\catname{SLat}} \otimes I_{\catname{SLat}}))\\
            &\cong \coprod_{A} (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A \times B} I_{\catname{SLat}}\\
            &\cong F(A \times B)
      \end{align*}
    In the above we used the fact that functors $- \otimes X$ and $X \otimes -$ preserve colimits as they are both left-adjoint by symmetry and monoidal closedness of $\catname{SLat}$.
    \end{proof}

@@

      \[\adjustbox{width=\linewidth}{
      \begin{tikzcd}  
          & {f + g \otimes *} && {\triangle_{A} \otimes ((f+g) \times (f+g))} \\
          f + g &&&& {\triangle_{A'};((f + g) \times (f+g)) = (f + g);\triangle_{A'};} \\
          & {* \otimes f + g} && {f + g \otimes \triangle_{A'}}
          \arrow[from=1-2, to=1-4]
          \arrow[from=1-4, to=2-5]
          \arrow[from=2-1, to=1-2]
          \arrow[from=2-1, to=3-2]
          \arrow[from=3-2, to=3-4]
          \arrow[from=3-4, to=2-5]
      \end{tikzcd}}
      \]

@@

\subsection{Proofs for Section \ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:pushout}

@@

\subsection{Existence of pushout}

@@

    S_{V} &= \{\\
          &\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\

@@

          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\

@@

          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\\
          &\}\\

@@

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
\ifdefined \ONECOLUMN
\begin{align*}
    \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{
        &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
        &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\}    
\end{align*}
\else
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = (&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &<_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        &)    
    \end{align*}
\fi
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[]_{V}^{*}$ is its extension to sequences.
There is also $[] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

@@

    \begin{enumerate}
        \item The first case is when $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) = \varnothing$.
              Then the property trivially holds as the empty set is a subset of any set.
        \item Suppose $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$.
              Then, 
              \[
                \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])) = [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))
              \]
              since $v \sim v$,
              and hence
              \ifdefined \ONECOLUMN
              \begin{align*}
                [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))
                &\subseteq
                \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
                &\subseteq
                (\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])))^{c} = \consistency_{\sfrac{X+Y}{\sim}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
              \end{align*}
              \else
              \begin{align*}
                &[[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))\\
                &\subseteq\\
                &\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
                &\subseteq\\
                &(\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])))^{c} = \consistency_{\sfrac{X+Y}{\sim}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
              \end{align*}
              \fi
              by recalling that $[[]_{V}^{\consistency}, []_{E}^{\consistency}] = [[]_{V};\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}, []_{E};\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}]$.
    \end{enumerate}

@@

@@

\subsection{Proofs for Section \ref{sec:soundness-and-completeness}: Soundness and Completeness}
\label{sec:appendix:detailed_proofs}
In this section we will elaborate on soundness and completeness of our translation $\llbracket - \rrbracket : \catname{S}^{+}(\Sigma, \mathcal{E}) \to \WellTypedMdaEcospans / \mathcal{S}, \mathcal{E}$.
First we recite the interpretation function $[]$ on $\Sigma$-terms as defined in~\cite{bonchi_string_2022-2} that induces a functor $\catname{S}(\Sigma) \to \MdaCospans$ and eventually a functor $\catname{S}^{+}(\Sigma) \to \MdaCospans^{+}$.
The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Figure~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

@@

@@

@@

            n_2 + m_1 \xrightarrow{} n_2' \setminus (n_1' \setminus n_1) + m_1 \xrightarrow{} \mathcal{L}_{2}^{\bot} \xleftarrow{} m'_2 \setminus (m_1' \setminus m_1) + n_1 \xleftarrow{} m_2 + n_1    
            \] is an MDA cospan.

@@

@@

@@

\subsection{Additional figures for Section~\ref{sec:introduction}:Introduction}

@@

\subsection{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:iso}

@@

\subsection{DPO rewriting for arbitrary signatures}
\label{sec:dpo-fix}
Here we will discuss how one can formulate DPO rewrite rules where the left-hand of right-hand sides contain sub-hypergraphs with no inputs and outputs at the outermost level.
First we will illustrate where the restriction in the original formulation of the DPO rewriting as per the Definition~\ref{def:convex_dpo} comes from via several examples: one for Cartesian equations (which are important to express e-graphs as e-hypergraphs) and two for the structural rules we need for $\WellTypedMdaEcospans / \mathcal{S}$ to be a category enriched over $\catname{SLat}$.
Consider a rewrite rule induced by the Cartesian structure $a;! = id_{I}$ for $a : 0 \to 1$ which is rendered string diagrammatically as 
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/cartesian_rewrite_string}
}
\]
which becomes the following extended cospan of e-hypergraphs
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/cartesian_rewrite_hypergraph}
}
\]

@@

\begin{figure*}
    \begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.7}{
    \tikzfig{../figures/appendix/cartesian_rewrite_source_graph}
    }
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.55\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/cartesian_rewrite_dpo_stuck}
        }
        \]
    \end{subfigure}
    \caption{DPO rewriting stuck}
    \label{fig:dpo-stuck}
\end{figure*}
Another problematic rewrite arises when we want to utilise the structural rule of the free enrichment
\[
\adjustbox{width=\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
if we consider the following DPO square
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]

@@

We will define the action of $+$ on two cospans where either of the carriers is the empty e-hypergraph as

\begin{align*}
0 \xrightarrow{} n \xrightarrow{} &\;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
&\;+ \hspace{6em} \Coloneqq \hspace{2em} 0  \xrightarrow{} n \xrightarrow{} \;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
0 \xrightarrow{} 0 \xrightarrow{} &\;\varnothing \xleftarrow{} 0 \xleftarrow{} 0.
\end{align*}

\end{definition}

@@

\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_problem_fix}
}
\]

@@

\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]

@@

\begin{figure}[h!]
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_elimination_hypergraph}
    }
    \]
\end{subfigure}
\hfill
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_introduction_hypergraph}
    }
    \]
\end{subfigure}
\end{figure}

and similar rules to introduce and eliminate in nested contexts (Figure~\ref{fig:nested_elimination}).

@@

\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_fix}
}
\]

and the full DPO square becomes

\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_dpo_fix}
}
\]

The target e-hypergraph is not exactly what we wanted, however, but we can use the auxiliary rule to remove this unlabelled edge.

>>>>

\ifdefined\ONECOLUMN
% \section{}
\section{$\catname{SLat}$}
\else
\subsection{$\catname{SLat}$}
\fi
\label{sec:appendix:slat}

@@

    Note that this construction does not identify $(s_1,s_2) +_{S_1 \otimes S_2} (s_1',s_2')$ and $(s_1 +_{S_1} s_1', s_2 +_{S_2} s_2')$.
    The unit for this tensor product is $I = \{*\}$ --- a one-element semilattice.
    Clearly $S \otimes I \cong S$ by mapping $(s,*) \mapsto s$ for all $s \in S$ and vice versa.
    The symmetry, associators and unitors are then obvious morphisms.
    Finally, the category is closed since the set of homomorpisms between two semilattices is a semilattice by defining $(f + g)(x)$ as $f(x) + g(x)$.
    $f + g$ is a homomorphism since $(f + g)(x+y) = f(x+y) + g(x+y) = f(x) + f(y) + g(x) + g(y) = f(x) + g(x) + f(y) + g(y) = f(x+y) + g(x+y)$.
  \end{proof}

  This makes $\catname{SLat}$ a suitable base for enrichment.
%    and allows us to take $\mathcal{V} = \catname{SLat}$.

@@

    \begin{proposition}[Special case of Proposition 6.4.6~\cite{Borceux_1994}]
      The forgetful functor $U : \catname{SLat} \to \catname{Set}$ has a left adjoint free functor $F : \catname{Set} \to \catname{SLat}$.
    \end{proposition}
    \begin{proof}
      The functor $F$ is defined by letting $F(A) = \coprod_{A} I_{\catname{SLat}}$ where the latter is a coproduct which is a `free' product of semilattices defined as follows.
      The elements of $S_{1} \coprod S_{2}$ are sequences $s_{1} + s_{2} + \ldots + s_{n}$ where each $s_{i}$ is either from $S_{1}$ or $S_{2}$ quotiented by all relations in $S_{1}$ and $S_{2}$ and by the necessary equations to turn it into a semilattice.
      The adjunction is then given by the following natural isomorphisms
      \begin{align*}
      \catname{SLat}(\coprod_{A}(I_{\catname{SLat}}), B) &\cong \prod_{A}(\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A,\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A, U(B))
      \end{align*}
      The first isomorphism is given by the fact that hom-functor makes limits into colimits in its first argument, the second being given by the property that $|\catname{Set}(A,B)| = |B|^{|A|} = |\underbrace{B \times \ldots \times B}_{|A|}|$, and the last isomorphism is given by the definition of $U$.
      Furthermore, we have, 
      \[
      F(I) \cong I_{\catname{SLat}}
      \]
      and 
      \begin{align*}
      F(A) \otimes F(B) &\cong (\coprod_{A} I_{\catname{SLat}}) \otimes (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (I_{\catname{SLat}} \otimes \coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (\coprod_{B} (I_{\catname{SLat}} \otimes I_{\catname{SLat}}))\\
            &\cong \coprod_{A} (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A \times B} I_{\catname{SLat}}\\
            &\cong F(A \times B)
      \end{align*}
    In the above we used the fact that functors $- \otimes X$ and $X \otimes -$ preserve colimits as they are both left-adjoint by symmetry and monoidal closedness of $\catname{SLat}$.
    \end{proof}

@@

      \[\adjustbox{width=\linewidth}{
      \begin{tikzcd}  
          & {f + g \otimes *} && {\triangle_{A} \otimes ((f+g) \times (f+g))}\\
          f + g &&&& {\triangle_{A'};((f + g) \times (f+g)) = (f + g);\triangle_{A'}}\\
          & {* \otimes f + g} && {f + g \otimes \triangle_{A'}}
          \arrow[from=1-2, to=1-4]
          \arrow[from=1-4, to=2-5]
          \arrow[from=2-1, to=1-2]
          \arrow[from=2-1, to=3-2]
          \arrow[from=3-2, to=3-4]
          \arrow[from=3-4, to=2-5]
      \end{tikzcd}}
      \]

@@

\ifdefined\ONECOLUMN
\section{Proofs for Section \ref{sec:e-hypergraphs}: E-hypergraphs}
\else
\subsection{Proofs for Section \ref{sec:e-hypergraphs}: E-hypergraphs}
\fi
\label{sec:appendix:pushout}

@@

\ifdefined\ONECOLUMN
\section{Existence of pushout}
\else
\subsection{Existence of pushout}
\fi

@@

    S_{V} = \{&
          \;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\

@@

          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\}\\
          \cup&\\
          \{&
          \;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\

@@

          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\}\\
          \cup&\\
          \{
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\}\\

@@

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
\ifdefined \ONECOLUMN
\begin{align*}
    \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{
        &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
        &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\}    
\end{align*}
\else
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = (&
            \sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &<_{\sfrac{X + Y}{\sim (R_{V},R_{E})}})    
    \end{align*}
\fi
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[]_{V}^{*}$ is its extension to sequences.
There is also $[] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

@@

    \begin{enumerate}
        \item The first case is when $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) = \varnothing$.
              Then the property trivially holds as the empty set is a subset of any set.
        \item Suppose $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$.
              Then, 
              \[
                \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])) = [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))
              \]
              since $v \sim v$,
              and hence
              \ifdefined \ONECOLUMN
              \begin{align*}
                [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))
                &\subseteq
                \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
                &\subseteq
                (\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])))^{c} = \consistency_{\sfrac{X+Y}{\sim}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
              \end{align*}
              \else
              \begin{align*}
                &[[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}})(v))\\
                &\subseteq\\
                &\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
                &\subseteq\\
                &(\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])))^{c} = \consistency_{\sfrac{X+Y}{\sim}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))\\
              \end{align*}
              \fi
              by recalling that 
              \[
              [[]_{V}^{\consistency}, []_{E}^{\consistency}] = [[]_{V};\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}, []_{E};\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}]
              \].
    \end{enumerate}

@@

\begin{lemma}
$\consistency_{\sfrac{X+Y}{\sim}}^{p}$ is not total.
\end{lemma}
\begin{proof}
The property holds as we require that the arrows of the span for the pushout do not map vertices to different components and therefore they can not be merged into one component.
\end{proof}

@@

\ifdefined\ONECOLUMN
\section{Proofs for Section \ref{sec:soundness-and-completeness}: Soundness and Completeness}
\else
\subsection{Proofs for Section \ref{sec:soundness-and-completeness}: Soundness and Completeness}
\fi
\label{sec:appendix:detailed_proofs}
In this section we will elaborate on soundness and completeness of our translation $\llbracket - \rrbracket : \catname{S}^{+}(\Sigma, \mathcal{E}) \to \WellTypedMdaEcospans / \mathcal{S}, \mathcal{E}$.
First we recite the interpretation function $[]$ on $\Sigma$-terms as defined in~\cite{bonchi_string_2022-2} that induces a functor $\catname{S}(\Sigma) \to \MdaCospans$ and eventually a functor $\catname{S}^{+}(\Sigma) \to \MdaCospans^{+}$.
The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Figure~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

@@

              \adjustbox{scale=0.8}{

@@

              \adjustbox{width=0.7\linewidth}{

@@

            % fix overflowing
            % n_2 + m_1 \xrightarrow{} n_2' \setminus (n_1' \setminus n_1) + m_1 \xrightarrow{} \mathcal{L}_{2}^{\bot} \xleftarrow{} m'_2 \setminus (m_1' \setminus m_1) + n_1 \xleftarrow{} m_2 + n_1    
            \tikzcdset{row sep/mysizerow/.initial=0.1ex}
            \tikzcdset{column sep/mysizecol/.initial=0.1ex}
            \[\
\begin{tikzcd}[cramped, row sep=mysizerow, column sep = mysizecol]
                n_2 + m_1 &&&&&&&& m_2 + n_1\\
                & \hspace{-2em} n_2' \setminus (n_1' \setminus n_1) + m_1 &\; & & \mathcal{L}_{2}^{\bot} &\; && m'_2 \setminus (m_1' \setminus m_1) + n_1\hspace{-2em} &\\
                \arrow[from=1-1, to=2-2]
                \arrow[from=1-9, to=2-8]
                \arrow[from=2-2, to=2-5]
                \arrow[from=2-8, to=2-5]
            is an MDA cospan.

@@

    n \xrightarrow{} n'_{1} \xrightarrow{} \mathcal{C}_{1} \xleftarrow{} m'_{1} + i + k \xleftarrow{} i+k \xrightarrow{} i' + k \xrightarrow{} \mathcal{L} \xleftarrow{} k + j' \xleftarrow{} k + j \xrightarrow{} n'_2 + k + j \xrightarrow{} \mathcal{C}_{2} \xleftarrow{} m'_2 \xleftarrow{} m

@@

        n \to n_{1}' \cup n_{2}' \cup i' \to \mathcal{C}_{1} \cup \mathcal{C}_{2} \cup \mathcal{L} \xleftarrow{} m_{1}' \cup m_{2}' \cup j' \xleftarrow{} m  = n \xrightarrow{f} n' \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} m' \xleftarrow{g} m

@@

\ifdefined\ONECOLUMN
\section{Additional figures for Section~\ref{sec:introduction}:Introduction}
\else
\subsection{Additional figures for Section~\ref{sec:introduction}:Introduction}
\fi

@@

\ifdefined\ONECOLUMN
\section{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\else
\subsection{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\fi
\label{sec:appendix:iso}

@@

\ifdefined\ONECOLUMN
\section{DPO rewriting for arbitrary signatures}
\else
\subsection{DPO rewriting for arbitrary signatures}
\fi
\label{sec:dpo-fix}
Here we will discuss how one can formulate DPO rewrite rules where the left-hand of right-hand sides contain sub-hypergraphs with no inputs and outputs at the outermost level.
First we will illustrate where the restriction in the original formulation of the DPO rewriting as per the Definition~\ref{def:convex_dpo} comes from via several examples: one for Cartesian equations (which are important to express e-graphs as e-hypergraphs) and two for the structural rules we need for $\WellTypedMdaEcospans / \mathcal{S}$ to be a category enriched over $\catname{SLat}$.
Consider a rewrite rule induced by the Cartesian structure $a;! = id_{I}$ for $a : 0 \to 1$ which is rendered string diagrammatically as 
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/cartesian_rewrite_string}
}
\]
which becomes the following extended cospan of e-hypergraphs
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/cartesian_rewrite_hypergraph}
}
\]

@@

\begin{figure*}
    \begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.7}{
    \tikzfig{../figures/appendix/cartesian_rewrite_source_graph}
    }
    \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.55\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/cartesian_rewrite_dpo_stuck}
        }
        \]
    \end{subfigure}
    \caption{DPO rewriting stuck}
    \label{fig:dpo-stuck}
\end{figure*}
Another problematic rewrite arises when we want to utilise the structural rule of the free enrichment
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.6\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
\tikzfig{../figures/appendix/structural_rewrite_problem}
}
\]
\fi
if we consider the following DPO square
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck}
}
\]
\fi

@@

We will define the action of $+$ on two cospans where either of the carriers is the empty e-hypergraph as
\begin{align*}
0 \xrightarrow{} n \xrightarrow{} &\;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
&\;+ \hspace{6em} \Coloneqq \hspace{2em} 0  \xrightarrow{} n \xrightarrow{} \;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
0 \xrightarrow{} 0 \xrightarrow{} &\;\varnothing \xleftarrow{} 0 \xleftarrow{} 0.
\end{align*}
\end{definition}

@@

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_problem_fix}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_problem_fix}
}
\]
\fi

@@

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\else
\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/structural_rewrite_dpo_stuck_fix}
}
\]
\fi

@@

\ifdefined\ONECOLUMN
\begin{figure}[h!]
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_elimination_hypergraph}
        }
        \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
        \adjustbox{scale=0.7}{
        \tikzfig{../figures/appendix/idI_introduction_hypergraph}
        }
        \]
    \end{subfigure}
    \end{figure}
\else
\begin{figure}[h!]
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_elimination_hypergraph}
    }
    \]
\end{subfigure}
\hfill
\begin{subfigure}{0.45\linewidth}
    \[
    \adjustbox{scale=0.55}{
    \tikzfig{../figures/appendix/idI_introduction_hypergraph}
    }
    \]
\end{subfigure}
\end{figure}
\fi
and similar rules to introduce and eliminate in nested contexts (Figure~\ref{fig:nested_elimination});

@@

\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.6\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_fix}
}
\]
\else
\[
\adjustbox{width=0.8\linewidth}{
    \tikzfig{../figures/appendix/cartesian_rewrite_fix}
}
\]
\fi
and the full DPO square becomes
